Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-23 main.tex
working in X

\section{Main Results}

@@

\begin{thm}\label{thm:CP-DI}
  Let $A$ be a rectified formula, and assume we have a combinatorial proof $\phi\colon\gC\to\graphof A$. Then there is a derivation
  \begin{equation}
    \label{eq:decom}
    \vlderivation{
      \vlde{\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{\Deri_2}{A}{
        \vlde{\FOMLS}{\Deri_1}{A'}{
          \vlhy{\ttt}}}}
  \end{equation}
  for some $A'$ such that $\graphof{\rectif{A'}}=\gC$. Conversely,
  whenever we have a derivation as in~\eqref{eq:decom} above, then
  there is a combinatorial proof $\phi\colon\gC\to\graphof A$ such that $\gC=\graphof{\rectif{A'}}$.
\end{thm}

@@

\section{Translating between $\FOLK$ and $\FOKS$}

@@

\section{Fonets and Linear Proofs}

@@

Now we are going to show how from a given fonet
$\tuple{\gC,\linkingof\gC}$ we can construct a sequent proof $\Pi$ in
$\FOMLL$ such that $\fographof\Pi=\tuple{\gC,\linkingof\gC}$. In the
proof net literature, this operation is also called
\emph{sequentialization}. The basic idea behind our sequentialization
is to construct a propositional linked cograph, called the \bfit{frame}~\cite{hughes:unifn}
of $\gC$, in which the dependencies between the binders in $\gC$ are
encoded as links. Then we can apply the \emph{splitting tensor
theorem} to the frame, and then reconstruct the sequent proof $\Pi$.

@@

\section{Skew Bifibrations and Resource Management}

@@

\begin{lemma}
  Let $\downsmash{\vlderivation{
    \vlde{\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{\Deri}{B}{
      \vlhy{A}}}}$ be a derivation. Then there is a recified derivation
  $\vlderivation{
    \vlde{\set{\wrD,\rectif\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\rectif\Deri}{\rectif
      B}{ \vlhy{\rectif A}}}$, such that the induced maps
  $\mapof\Deri\colon\graphof{A}\to\graphof{B}$ and
  $\mapof{\rectif\Deri}\colon\graphof{\rectif A}\to\graphof{\rectif
    B}$ are equal up to a variable renaming of the vertex labels.
\end{lemma}

@@

\begin{lemma}\label{skew->cw}
  Let $\gA$ and $\gB$ be fographs, let $\phi\colon\gA\to\gB$ be a skew
  bifibration, and let $A$ and $B$ be formulas with $\graphof A=\gA$
  and $\graphof B=\gB$. Then there are derivations
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\acD,\rectif\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\rectif\Deri}{B}{
        \vlhy{A}}}
    \quand
    \vlderivation{
      \vlde{\set{\wrD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}}{\Deri}{B}{
        \vlhy{A\substof\phi}}}
  \end{equation*}
  such that $\mapof{\rectif\Deri}=\phi$ and $\rectif{\Deri}$ is a
  rectification of $\Deri$, and $\substof\phi$ is the substitution
  induced by $\phi$.
\end{lemma}

@@

\begin{proof}[Proof of Lemma~\ref{skew->cw}]
  By Lemma~\ref{lem:PE} we have $\gA=\graphof{\PE A}$ and
  $\gB=\graphof{\PE B}$.  Let $\vB'\subseteq\vB$ be the image of
  $\phi$, and let $\gB_1$ be the subgraph of $\gB$ induced by
  $\vB'$. Hence, we have two maps $\phi''\colon\gA\to\gB_1$ being a
  surjection and $\phi'\colon \gB_1\to\gB$ being an injection that
  reflects edges. \juihsuan{what do you mean by "reflect edges"?}\lutz{edge downstairs implies edge upstairs} Both, $\phi'$ and $\phi''$ remain skew
  bifibrations. Let us first look at $\phi'$.  Let $\PEp B_1$ be the
  propositional formula obtained from $\PE B$ by removing all atoms
  that are not represented by vertices in $\vB'$. Then $\graphof{\PEp
    B_1}=\gB_1$.  By \cite[Proposition~7.6.1]{str:07:RTA}, we have a
  derivation $\vlder{\set{\wrD,\fequ}}{\PE\Deri_1}{\PE B}{\PEp B_1}$.
  A subformula of $\PE B$ is called \emph{weak} if it has been
  introduced by a weakening. All subformulas of a weak formula are
  also weak. Two weak subformulas $B'$ and $B''$ of $\PE B$ form a
  \emph{weak pair} if $\PE B\fequ\Scons{B'\vlor B''}$ for some context
  $\Sconhole$. We can assume without loss of generality that
  whenever weak subformulas $B'$ and $B''$ form a weak pair, they have
  been introduced by the same instance of $\wrD$ in
  $\PE\Deri_1$.\footnote{If $\PE\Deri_1$ is not of that shape, it can
  brought into this form by simple rule permutations, and then
  collapsing several weakenings into a single one.}  Now we show that
  $\PE\Deri_1$ can be lifted. For this, observe that whenever a
  weakening in $\PE\Deri_1$ deletes an atom $x\in\VAR$, it must also
  delete all atoms in the scope of the corresponding quantifier,
  because $\phi'$ is a fibration on the binding graph. Hence, each line
  in $\PE\Deri_1$ is the propositional encoding $\PE P$ of a
  first-order formula $P$.  We now have to show
  that each instance of $\wrD$ is indeed a correct
  application in first-order logic. For this we have to inspect the
  cases a weakening happens inside a subformula $x\vlor C$ or $x\vlan
  C$ in $\PE\Deri_1$. There are the following cases:
  \begin{equation*}%\small
    \vlinf{}{}{S\cons{x\vlor D\vlor C}}{S\cons{x\vlor C}}
    \qquad
    \vlinf{}{}{S\cons{x\vlan (D\vlor C)}}{S\cons{x\vlan C}}
    \qquad
    \vlinf{}{}{S\cons{(x\vlor D)\vlan C}}{S\cons{x\vlan C}}
  \end{equation*}
  In the first case the weakening happens inside the scope of a
  $\forall$-quantifier, and in the second case inside the scope of a
  $\exists$-quantifier. Both cases are unproblematic in first-order
  logic. However, in the third case an $\exists$-quantifier would be
  transformed into an $\forall$-quantifier. But as $\phi$ has to
  preserve existentials, this third case cannot occur.
  Thus we have a first order derivation
  $\vlder{\set{\wrD,\fequ}}{\Deri_1}{B}{B_1}$ with $\PE B_1=\PEp B_1$.

@@

\section{Summary and Proofs of Main Results}

>>>>

\section{Main Results}
\label{sec:main}

@@

\begin{thm}\label{thm:CP-DI}
  Let  $\phi\colon\gC\to\gA$ be a combinatorial proof and let $A$ be a formula with $\gA=\graphof A$. Then there is a derivation
  \begin{equation}
    \label{eq:decom}
    \vlderivation{
      \vlde{\set{\wrD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}}{\Deri_2}{A}{
        \vlde{\FOMLS}{\Deri_1}{A'}{
          \vlhy{\ttt}}}}
  \end{equation}
  for some $A'\fequ C\substof\phi$ where $C$ is a formula with $\graphof
  C=\gC$ and $\substof\phi$ is the variable renaming substitution induced by~$\phi$.  Conversely,
  whenever we have a derivation as in~\eqref{eq:decom} above, then
  there is a combinatorial proof $\phi\colon\gC\to\graphof A$ such
  that $\gC=\graphof{\rectif{A'}}$.
\end{thm}

@@

\section{Translating between $\FOLK$ and $\FOKS$}
\label{sec:LK1-KS1}

@@

\section{Fonets and Linear Proofs}
\label{sec:linear}

@@

Now we are going to show how from a given fonet
$\tuple{\gC,\linkingof\gC}$ we can construct a sequent proof $\Pi$ in
$\FOMLL$ such that $\fographof\Pi=\tuple{\gC,\linkingof\gC}$. In the
proof net literature, this operation is also called
\emph{sequentialization}. The basic idea behind our sequentialization
is to construct a propositional linked cograph, called the \bfit{frame}~\cite{hughes:unifn}
of $\gC$, in which the dependencies between the binders in $\gC$ are
encoded as links. Then we can apply the \emph{splitting tensor
theorem} to the frame, and then reconstruct the sequent proof $\Pi$.
\lutz{if the proof of thm \ref{thm:MLS1->fonet} is verified, we can delete the frame-def here}

@@

\section{Skew Bifibrations and Resource Management}
\label{sec:skew}

@@

\begin{lemma}\label{lem:cw->rectif}
  Let $\downsmash{\vlderivation{
    \vlde{\set{\wrD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}}{\Deri}{B}{
      \vlhy{A}}}}$ be a derivation. Then there is a recified derivation
  $\vlderivation{
    \vlde{\set{\wrD,\rectif\acD,\rectif\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\rectif\Deri}{\rectif
      B}{ \vlhy{\rectif A}}}$, such that the induced maps
  $\mapof\Deri\colon\graphof{A}\to\graphof{B}$ and
  $\mapof{\rectif\Deri}\colon\graphof{\rectif A}\to\graphof{\rectif
    B}$ are equal up to a variable renaming of the vertex labels.
\end{lemma}

@@

\begin{lemma}\label{lem:skew->cw}
  Let $\gA$ and $\gB$ be fographs, let $\phi\colon\gA\to\gB$ be a skew
  bifibration, and let $A$ and $B$ be formulas with $\graphof A=\gA$
  and $\graphof B=\gB$. Then there are derivations
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\acD,\rectif\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\rectif\Deri}{B}{
        \vlhy{A}}}
    \quand
    \vlderivation{
      \vlde{\set{\wrD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}}{\Deri}{B}{
        \vlhy{A\substof\phi}}}
  \end{equation*}
  such that $\mapof{\rectif\Deri}=\phi$ and $\rectif{\Deri}$ is a
  rectification of $\Deri$, and $\substof\phi$ is the substitution
  induced by $\phi$.
\end{lemma}

@@

\begin{proof}[Proof of Lemma~\ref{lem:skew->cw}]
  By Lemma~\ref{lem:PE} we have $\gA=\graphof{\PE A}$ and
  $\gB=\graphof{\PE B}$.  Let $\vB'\subseteq\vB$ be the image of
  $\phi$, and let $\gB_1$ be the subgraph of $\gB$ induced by
  $\vB'$. Hence, we have two maps $\phi''\colon\gA\to\gB_1$ being a
  surjection and $\phi'\colon \gB_1\to\gB$ being an injection that
  reflects edges. \juihsuan{what do you mean by "reflect edges"?}\lutz{edge downstairs implies edge upstairs} Both, $\phi'$ and $\phi''$ remain skew
  bifibrations. Let us first look at $\phi'$.  Let $\PEp B_1$ be the
  propositional formula obtained from $\PE B$ by removing all atoms
  that are not represented by vertices in $\vB'$. Then $\graphof{\PEp
    B_1}=\gB_1$.  By \cite[Proposition~7.6.1]{str:07:RTA}, we have a
  derivation $\vlder{\set{\wrD,\fequ}}{\PE\Deri_1}{\PE B}{\PEp B_1}$.
  A subformula of $\PE B$ is called \emph{weak} if it has been
  introduced by a weakening. All subformulas of a weak formula are
  also weak. Two weak subformulas $B'$ and $B''$ of $\PE B$ form a
  \emph{weak pair} if $\PE B\fequ\Scons{B'\vlor B''}$ for some context
  $\Sconhole$. We can assume without loss of generality that
  whenever weak subformulas $B'$ and $B''$ form a weak pair, they have
  been introduced by the same instance of $\wrD$ in
  $\PE\Deri_1$.\footnote{If $\PE\Deri_1$ is not of that shape, it can
  brought into this form by simple rule permutations, and then
  collapsing several weakenings into a single one.}  Now we show that
  $\PE\Deri_1$ can be lifted. For this, observe that whenever a
  weakening in $\PE\Deri_1$ deletes an atom $x\in\VAR$, it must also
  delete all atoms in the scope of the corresponding quantifier,
  because $\phi'$ is a fibration on the binding graph. Hence, each line
  in $\PE\Deri_1$ is the propositional encoding $\PE P$ of a
  first-order formula $P$.  We now have to show
  that each instance of $\wrD$ is indeed a correct
  application in first-order logic. For this we have to inspect the
  cases a weakening happens inside a subformula $x\vlor C$ or $x\vlan
  C$ in $\PE\Deri_1$. There are the following cases:
  \begin{equation*}%\small
    \vlinf{}{}{S\cons{x\vlor D\vlor C}}{S\cons{x\vlor C}}
    \qquad
    \vlinf{}{}{S\cons{x\vlan (D\vlor C)}}{S\cons{x\vlan C}}
    \qquad
    \vlinf{}{}{S\cons{(x\vlor D)\vlan C}}{S\cons{x\vlan C}}
  \end{equation*}
  In the first case the weakening happens inside the scope of a
  $\forall$-quantifier, and in the second case inside the scope of a
  $\exists$-quantifier. Both cases are unproblematic in first-order
  logic. However, in the third case an $\exists$-quantifier would be
  transformed into an $\forall$-quantifier. But as $\phi$ has to
  preserve existentials, this third case cannot occur.
  Thus we have a first order derivation
  $\vlder{\set{\wrD,\fequ}}{\Deri_1}{B}{B_1}$ with $\PE B_1=\PEp B_1$.

@@

\section{Summary and Proof of Main Result}
\label{sec:summary}

The only theorem of Section~\ref{sec:main} that has not yet been
proved is Theorem~\ref{thm:CP-DI} establishing the full correspondence
between decomposed proofs in $\FOKS$ and combinatorial proofs. We show
the proof here, by summarizing the results of the previous two
Sections~\ref{sec:linear} and~\ref{sec:skew}.

\begin{proof}[Proof of Theorem~\ref{thm:CP-DI}]
  First, asssume we have a combinatorial proof $\phi\colon\gC\to\gA$
  be a combinatorial proof and a formula $A$ with $\gA=\graphof A$.
  Let $C$ be a formula with $\graphof C=\gC$, and let $\substof\phi$
  be the substitution induced by $\phi$.
  By Lemma~\ref{lem:skew->cw} there is a derivation
  \begin{equation*}
    \vlder{\set{\wrD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}}{\Deri_2}{A}{C\substof\phi}
  \end{equation*}
  Since $\gC$ is a fonet, we have by Theorem~\ref{thm:fonet->MLS1} a derivation
  \begin{equation*}
    \vlder{\FOMLS}{\Deri'_1}{C}{\ttt}
  \end{equation*}
  This derivation remains valid if we apply the substitution
  $\substof\phi$ to every line in $\Deri'_1$, yielding the derivation
  $\Deri_1$ of $C\substof\phi$ as desired.

  Conversely, assume we have a decomposed derivation
  \begin{equation}
    \label{eq:decom}
    \vlderivation{
      \vlde{\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{\Deri_2}{A}{
        \vlde{\FOMLS}{\Deri_1}{A'}{
          \vlhy{\ttt}}}}
  \end{equation}
  Then we can transform $\Deri_1$ into a recified form
  $\rectif\Deri_1$, proving $\rectif A'$. By
  Theorem~\ref{thm:MLS1->fonet}, the linked fograph
  $\fographof{\rectif\Deri_1}=\tuple{\fographof{\rectif
      A'},\linkingof{\rectif\Deri_1}}$ is a fonet.  Then, by
  Lemma~\ref{lem:cw->rectif}, there is a rectified derivation
  $\vlderivation{
    \vlde{\set{\wrD,\rectif\acD,\rectif\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\rectif{\Deri_2}}{\rectif
      A}{ \vlhy{\rectif{A'}}}}$ whose induced map
  $\mapof{\rectif{\Deri_2}}\colon\graphof{\rectif{A'}}\to\graphof{\rectif
    A}$ is the same a the induced map
  $\mapof{\Deri_2}\colon\graphof{A'}\to\graphof{A}$ of $\Deri_2$. By
  Lemma~\ref{lem:cw->skew}, this map is a skew bifibration. Hence, we have a combinatorial proof \lutz{shit, something's wrong...}

  %% Let $\downsmash{\vlderivation{
  %%   \vlde{\set{\wrD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}}{\Deri}{B}{
  %%     \vlhy{A}}}}$ be a derivation. Then there is a recified derivation
  %% $\vlderivation{
  %%   \vlde{\set{\wrD,\rectif\acD,\rectif\cfaD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\rectif\Deri}{\rectif
  %%     B}{ \vlhy{\rectif A}}}$, such that the induced maps
  %% $\mapof\Deri\colon\graphof{A}\to\graphof{B}$ and
  %% $\mapof{\rectif\Deri}\colon\graphof{\rectif A}\to\graphof{\rectif
  %%   B}$ are equal up to a variable renaming of the vertex labels.

  %% for some $A'\fequ C\sigma$ where $C$ is a formula with $\graphof
  %% C=\gC$ and $\sigma$ is a variable renaming substitution.  Conversely,
  %% whenever we have a derivation as in~\eqref{eq:decom} above, then
  %% there is a combinatorial proof $\phi\colon\gC\to\graphof A$ such
  %% that $\gC=\graphof{\rectif{A'}}$.
